---
title: "Point Spread Teaser Analysis"
author: "Cory Waters"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Load data

```{r load_data, message=FALSE}
nfl_games <- read_csv('data/nfl-book.csv') %>% 
  select(seas, wk, h, v, ptsh, ptsv, sprv, ou)
nfl_games <- filter(nfl_games, seas >= 2008)
```

## Create point spread results table

```{r}
nfl <- nfl_games %>% 
  mutate(mov = ptsh - ptsv,
         home_fav = as.integer(sprv >= 0),
         road_fav = as.integer(sprv < 0),
         home_cover = as.integer(mov > sprv),
         road_cover = as.integer(mov < sprv),
         fav_cover = as.integer((home_fav + home_cover == 2) | (road_fav + road_cover == 2)),
         dog_cover = as.integer((home_fav + road_cover == 2) | (road_fav + home_cover == 2)),
         push = as.integer(mov == sprv)) %>% 
  mutate(spread_group = case_when( # category of spreads crossing 3/7/10/14
    abs(sprv) >= 0 & abs(sprv) < 3 ~ "0 to 2.5",
    abs(sprv) >= 3 & abs(sprv) < 7 ~ "3 to 6.5",
    abs(sprv) >= 7 & abs(sprv) < 10 ~ "7 to 9.5",
    abs(sprv) >= 10 & abs(sprv) < 14 ~ "10 to 13.5",
    TRUE ~ "14+"
  )) 
```


## Visualize spreads

```{r}
nfl %>% 
  group_by(seas, spread_group) %>% 
  count() %>% 
  ggplot(aes(reorder(spread_group,n), n)) +
  geom_point() +
  coord_flip() +
  labs(x = 'Spread', y = 'Number Games') +
  facet_wrap(~seas,nrow = 4) +
  theme_minimal()
```

Distribution of point spreads

```{r}
nfl %>%
  group_by(spread_group, seas) %>%
  count() %>%
  ggplot(aes(reorder(spread_group,n), seas, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(x = 'point spread', y = 'season', fill = 'games') +
  theme(panel.grid = element_blank()) 
```

## Fav vs Dog 

```{r}
nfl %>% 
  summarise(tot_games = n(),
            fav = sum(fav_cover),
            dog = sum(dog_cover),
            push = sum(push)
            ) %>% 
  transmute(fav_pct=fav/(tot_games - push),
            dog_pct =dog/(tot_games - push))
```

## Spread results by grouping

```{r}
nfl %>% 
  group_by(spread_group) %>% 
  summarise(tot_games = n(),
            fav = sum(fav_cover),
            dog = sum(dog_cover),
            push = sum(push)
            ) %>% 
  ungroup() %>% 
  mutate(fav_pct = fav / (tot_games - push), # remove push
         dog_pct = dog / (tot_games - push)
         ) %>% 
  select(spread_group, fav_pct, dog_pct, tot_games)



```

## Home vs Away 

```{r}
# By Home
nfl %>% 
  summarise(home = sum(home_cover),
            away = sum(road_cover),
            push = sum(push)) %>% 
  mutate(tot_games = home + away) %>% 
  mutate(home_pct = home / tot_games,
         away_pct = away / tot_games) %>% 
  select(home_pct, away_pct)
```

## Home Fav

```{r}
nfl %>% 
  filter(push != 1) %>% # remove push
  group_by(home_fav) %>% 
  summarise(home = sum(home_cover),
            away = sum(road_cover)
            ) %>% 
  ungroup() %>% 
  mutate(tot_games = (home + away),
         h_pct = home / tot_games,
         a_pct = away / tot_games) %>% 
  select(home_fav, h_pct, a_pct)
```

## Reshape to fav/dog from home/away

```{r}
nfl_reshape <- nfl %>% 
  transmute(fav_team = ifelse(sprv >= 0, h, v),
         dog_team = ifelse(sprv < 0, h, v),
         favored = case_when(fav_team == h ~ 'H',
                             fav_team == v ~ 'A',
                             TRUE ~ 'E'),
         fav_score = ifelse(fav_team == h, ptsh, ptsv),
         dog_score = ifelse(fav_team == h, ptsv, ptsh),
         mov_fav = fav_score - dog_score,
         point_spread = abs(sprv),
         total = ou)
```

## Compare spreads to MOV

```{r}
nfl_reshape %>% 
  group_by(point_spread, mov_fav) %>% 
  count() %>% 
  ggplot(aes(point_spread, mov_fav,color=n)) +
  geom_hex() +
  scale_fill_distiller(direction = -1,palette = 7) +
  theme_minimal()
  
```

## View distribution of spreads and totals

```{r}
nfl_reshape %>% 
  select(point_spread,total) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  geom_density() +
  facet_wrap(~key, scales = 'free') +
  theme_minimal()
```

## Fav cover

```{r}
nfl_reshape %>% 
  filter(mov_fav != point_spread) %>% # remove pushes
  mutate(fav_cover = mov_fav > point_spread) %>% 
  group_by(fav_cover) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate(pct = n / sum(n))
```

## How many NFL games push

```{r}
summarise(nfl_reshape, push_pct = mean(mov_fav == point_spread))
```

## Calculate teaser hit rates

```{r}
buy_pts <- c(6,6.5,7,7.5,10,13,14,17,21) # number of points to add to sprd

map_df(
  buy_pts, # pass the teaser points vector
  ~ nfl_reshape %>%
    mutate(
      #fav_cover = mov_fav > point_spread,
      with_pts_fav = mov_fav > (point_spread - .x), # subtract pts from fav
      with_pts_dog = mov_fav < (point_spread + .x) # add points to dog
    ) %>%
    summarise(
      tot_games = n(), 
      tease_fav = sum(with_pts_fav),
      tease_dog = sum(with_pts_dog)
    ) %>%
    mutate(
      fav_wpct = tease_fav / tot_games,
      dog_wpct = tease_dog / tot_games
    ) %>%
    add_column(pts = .x, .before = 1) # add the point buys as col
) %>% 
  mutate(combined_pct = (tease_fav + tease_dog)/(tot_games * 2)) %>% 
  select(pts, fav_wpct, dog_wpct, combined_pct)

```

## Tease indivdual spreads

```{r}

teaser_summary_by_pts <- map_df(
  buy_pts,
  ~ nfl_reshape %>%
    
    mutate(
      #fav_cover = mov_fav > point_spread,
      with_pts_fav = mov_fav > (point_spread - .x),
      with_pts_dog = mov_fav < (point_spread + .x)
    ) %>%
    group_by(point_spread) %>%
    summarise(
      tot_games = n(),
      tease_fav = sum(with_pts_fav),
      tease_dog = sum(with_pts_dog)
    ) %>%
    add_column(pts = .x, .before = 1)
) %>%
  select(point_spread, tot_games, pts, tease_fav, tease_dog)


teaser_summary_by_pts
```

## Get teaser results for point spread ranges

```{r}
get_teaser_results <- function(df, .sprd_min, .sprd_max) {
 filter(df, between(point_spread, .sprd_min, .sprd_max)) %>% 
    group_by(pts) %>% 
    summarise(games = sum(tot_games), 
              tease_fav = sum(tease_fav), 
              tease_dog = sum(tease_dog)) %>% 
    ungroup() %>% 
    mutate(fav_pct = tease_fav / games,
           dog_pct = tease_dog / games,
           combined_pct = (tease_fav + tease_dog)/(games * 2))
}

# get_teaser_results(teaser_summary_by_pts, 1.5, 2.5)
```

