---
author: "Cory Waters"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

## Load Data

```{r}
# Using data from the nflscrapr package
# https://github.com/ryurko/nflscrapR-data
# nfl_files <-
#   dir(
#   "~/Dropbox/nfl/nflscrapR-data-master/data/season_team_stats/",
#   full.names = TRUE
#   )
# nfl_stats <- map(nfl_files, read_csv)
nfl_seas <- dir("~/Dropbox/nfl/nflscrapR-data-master/data/season_games/", full.names = TRUE)

nfl_games <- map(nfl_seas, read_csv)
```

## Check data

```{r}
get_shape <- function(.list_df) {
  get_df_shape <- function(df) {
  x <- dim(df)
  tibble(nrow = x[1],
  ncol = x[2])
  }
  map_df(.list_df, get_df_shape)
}

get_shape(nfl_games)
```

## Combine seasons into one dataframe

```{r}
nfl_games <- bind_rows(nfl_games)
```

## Check colnames

```{r}
as_tibble(names(nfl_games))
```

## Convert from home vs away to team vs opp

```{r}

nfl_long <- nfl_games %>%
  select(GameID, Season, away, home, homescore, awayscore) %>%
  gather(location, team, -GameID, -Season, -homescore, -awayscore)
  
  
nfl_games <-
  nfl_long %>%
  left_join(nfl_long %>%
  select(GameID, Season,  opp = team)) %>%
  filter(team != opp) %>%
  mutate(
  tm_pts = ifelse(location == 'away', awayscore, homescore),
  opp_pts = ifelse(location == 'home', awayscore, homescore)
  ) %>%
  select(-awayscore, -homescore)
```

## Tally wins and losses

```{r}
nfl_games <- nfl_games %>% 
  mutate(mov = tm_pts - opp_pts,
         result = case_when(
           mov > 0 ~ 'win',
           mov < 0 ~ 'loss',
           TRUE ~ 'push'),
         score = 1) %>% 
  spread(result, score, fill = 0)

```

## Create Season Summary

```{r}
nfl_sum <- nfl_games %>% 
  group_by(Season, team) %>% 
  summarise_at(.vars = vars(tm_pts, opp_pts, win, loss, push), sum) %>% 
  ungroup()
```

## Add win percentage (ties half wins)

```{r}
nfl_sum <- nfl_sum %>% 
  mutate(games_played = win + loss + push,
         win_pct = (win + (push * .5)) / games_played)
```

## Create Pythagorean Expectation function

[Read about Pythagorean Expectation](https://en.wikipedia.org/wiki/Pythagorean_expectation)

```{r}
pythag_expect <- function(.ps, .pa, .ex = 2){
  # .ps = points scored
  # .pa = points against
  # .exponent
  # common exponents: mlb = 2, nfl = 2.37, nba = 13.91
  .ps^.ex / (.ps^.ex + .pa^.ex)
}
```

## Calculate Pythagorean Expectation per team

[2.37 exponent taken from Mathletics by Waye Winston](https://www.amazon.com/Mathletics-Gamblers-Enthusiasts-Mathematics-Basketball/dp/0691154589)

```{r}
nfl_sum <- nfl_sum %>% 
  mutate(pythag = pythag_expect(tm_pts, opp_pts, 2.37))
```

## Visualization expecation vs actual results

```{r}
nfl_sum %>%
  ggplot(aes(win_pct, pythag)) +
  geom_hex(bins=10) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_viridis_c() +
  theme_minimal()
```

## Visualize results per team vs expectation (2017)

```{r}
nfl_sum %>% 
  select(Season, team, pythag, win_pct) %>% 
  filter(Season == 2017) %>% 
  gather(stat, pct, -Season, -team) %>% 
  ggplot(aes(reorder(team,pct),pct)) +
  geom_point(aes(color=stat)) + 
  coord_flip() +
  #facet_wrap(~Season) + 
  labs(x = 'Team', y = 'Pct') +
  scale_y_continuous(labels = scales::percent_format()) + 
  theme_minimal()
```
